---
title: "Combining Data"
author: "WB"
date: "11/20/2019"
output: html_document
editor_options:
  chunk_output_type: console
---

# Combining the daily values we want to run in our forecasting model

## Preliminaries

```{r, message=FALSE, warning = FALSE}
library(lubridate)
library(tidyr)
library(dplyr)
library(reshape2)
library(ggplot2)
library(openair)
library(PerformanceAnalytics)
library(stringr)
library(RColorBrewer)

```


## Buoy sensor data:

- (Mahalia can fill out here the process she went through to get daily means, why we didn't go with max's, doing the time lags, etc.)

```{r, message=FALSE, warning = FALSE}
##########--------missisquoi bay--------##########
mb17.raw <- read.csv("MB17_daily.csv", 
                 header = T)
mb17 <- mb17.raw %>% 
  select(-c(X)) %>% 
  mutate(date = as.POSIXct(strptime(date, format = "%Y-%m-%d", tz = "Etc/GMT-4")))

mb18.raw <- read.csv("MB18_daily.csv", 
                 header = T)
mb18 <- mb18.raw %>% 
  select(-c(X)) %>% 
  mutate(date = as.POSIXct(strptime(date, format = "%Y-%m-%d", tz = "Etc/GMT-4")))
##################################################

##########--------st. albans bay--------##########
ab17.raw <- read.csv("SA17_daily.csv", 
                 header = T)
ab17 <- ab17.raw %>% 
  select(-c(X)) %>% 
  mutate(date = as.POSIXct(strptime(date, format = "%Y-%m-%d", tz = "Etc/GMT-4")))

ab18.raw <- read.csv("SA18_daily.csv", 
                 header = T)
ab18 <- ab18.raw %>% 
  select(-c(X)) %>% 
  mutate(date = as.POSIXct(strptime(date, format = "%Y-%m-%d", tz = "Etc/GMT-4")))
##################################################

```


## Meterological data: 

- going to use St AB HOBO met station data for both bays and both years because of inconsistencies with the MB met station (see `QAQC_MET_2017and02018.Rmd` for analysis)
- air temperature (C)
- wind speed (m/s)
- wind direction (m/s)
- solar radiation (W/m2) 
- other notes: its important to use package `openair` when getting averages for wind speed and direction because it takes into account the wind vectors

```{r, message=FALSE, warning = FALSE}

met17.raw <- read.csv("QAQC_dailyStABmet2017.csv", 
                      header = T)

met17 <- met17.raw %>% 
  select(-c(X)) %>% 
  mutate(date = as.POSIXct(strptime(date, format = "%Y-%m-%d", tz = "Etc/GMT-4")))

met18.raw <- read.csv("QAQC_dailyStABmet2018.csv", 
                      header = T)

met18 <- met18.raw %>% 
  select(-c(X)) %>% 
  mutate(date = as.POSIXct(strptime(date, format = "%Y-%m-%d", tz = "Etc/GMT-4")))


```

## Discharge data: 

- discharge from USGS gages: Missisquoi River (USGS Station 04294000) and Jewett Brook (USGS Station 04292810)


```{r, message=FALSE, warning = FALSE}
##########--------missisquoi river--------##########
mriv17.raw <- read.csv("mriv17_daily.csv", 
                      header = T)
mriv17 <- mriv17.raw %>% 
  select(-c(X)) %>% 
  mutate(date = as.POSIXct(strptime(date, format = "%Y-%m-%d", tz = "Etc/GMT-4"))) %>% 
  rename(discharge = DISCHARGE.rate)


mriv18.raw <- read.csv("mriv18_daily.csv", 
                      header = T)
mriv18 <- mriv18.raw %>% 
  select(-c(X)) %>% 
  mutate(date = as.POSIXct(strptime(date, format = "%Y-%m-%d", tz = "Etc/GMT-4"))) %>% 
  rename(discharge = discharge.rate)

#just to check 
# plot(mriv17$date, mriv17$DISCHARGE.rate, type = "l")
# plot(mriv18$date, mriv18$discharge.rate, type = "l")
##################################################

##########--------jewett brook--------##########
jb17.raw <- read.csv("jb17_daily.csv", 
                      header = T)
jb17 <- jb17.raw %>% 
  select(-c(X)) %>% 
  mutate(date = as.POSIXct(strptime(date, format = "%Y-%m-%d", tz = "Etc/GMT-4"))) %>% 
  rename(discharge = DISCHARGE.rate)


jb18.raw <- read.csv("jb18_daily.csv", 
                      header = T)
jb18 <- jb18.raw %>% 
  select(-c(X)) %>% 
  mutate(date = as.POSIXct(strptime(date, format = "%Y-%m-%d", tz = "Etc/GMT-4"))) %>% 
  rename(discharge = discharge.rate)

#just to check
#plot(jb17$date, jb17$DISCHARGE.rate, type = "l")
#plot(jb18$date, jb18$discharge.rate, type = "l")
##################################################

```

# Combining into two data frames for each bay 


```{r, message=FALSE, warning = FALSE}
##########--------missisquoi bay--------##########
mbAll17 <- merge(mb17, met17, by = "date")
mbAll17 <- merge(mbAll17, mriv17, by = "date")
#sometimes when using merge it'll duplicate dates so i'm going to run the `distint` funciton just in case 
mbAll17 <- mbAll17 %>% 
  distinct(date, .keep_all= TRUE)
#write.csv(mbAll17, "mbAll17.csv")
summary(mbAll17[,-1])

mbAll18 <- merge(mb18, met18, by = "date")
mbAll18 <- merge(mbAll18, mriv18, by = "date")

#sometimes when using merge it'll duplicate dates so i'm going to run the `distint` funciton just in case 
mbAll18 <- mbAll18 %>% 
  distinct(date, .keep_all= TRUE)
#write.csv(mbAll18, "mbAll18.csv")
summary(mbAll18[,-1])
##################################################

##########--------st. albans bay--------##########
abAll17 <- merge(ab17, met17, by = "date")
abAll17 <- merge(abAll17, jb17, by = "date")
#sometimes when using merge it'll duplicate dates so i'm going to run the `distint` funciton just in case 
abAll17 <- abAll17 %>% 
  distinct(date, .keep_all= TRUE)
#write.csv(abAll17, "abAll17.csv")

abAll18 <- merge(ab18, met18, by = "date")
abAll18 <- merge(abAll18, jb18, by = "date")
#sometimes when using merge it'll duplicate dates so i'm going to run the `distint` funciton just in case 
abAll18 <- abAll18 %>% 
  distinct(date, .keep_all= TRUE)
#write.csv(abAll18, "abAll18.csv")
##################################################

parameters <- colnames(mbAll17)
print(parameters)
```



# Basic visualizations for these data 


```{r, message=FALSE, warning = FALSE}
#first get data into long format

#missisquoi bay 
mbAll17_long <- mbAll17 %>% 
  mutate(year = year(date), 
         site = rep("MB")) %>% 
  gather(key = "parameter", value = "value", c(TempMean:discharge))

mbAll18_long <- mbAll18 %>% 
  mutate(year = year(date), 
         site = rep("MB")) %>% 
  gather(key = "parameter", value = "value", c(TempMean:discharge))

mbAll <- rbind(mbAll17_long, mbAll18_long)
mbAll$doy <- strftime(mbAll$date, format = "%j", tz = "Etc/GMT-4")
mb <- c("#FB9A99", "#E31A1C")


##########--------time series --------##########
mbAll %>% 
  filter(parameter == "PCMean1" | 
           parameter == "SpCondMean" | 
           parameter == "pHMean" | 
           parameter == "ODOMean" | 
           parameter == "ChlMean" | 
           parameter == "DeltaTempMean" |
           parameter == "ws" | 
           parameter == "solar" | 
           parameter == "temp" | 
           parameter == "discharge") %>% 
  ggplot(aes(x = doy, 
             y = value, 
             color = as.factor(year),
             group = as.factor(year))) + 
  geom_line()+
  scale_color_manual(values = mb)+
  theme_bw()+
  theme(text = element_text(size=14),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank())+
  facet_wrap(~parameter, scales = "free_y") +
  scale_x_discrete(breaks = seq(120,310,30)) +
  xlab("Day of year")+
  labs(color = "Year")
##################################################


##########--------boxplots--------##########
mbAll %>% 
  filter(parameter == "PCMean1" | 
           parameter == "SpCondMean" | 
           parameter == "pHMean" | 
           parameter == "ODOMean" | 
           parameter == "ChlMean" | 
           parameter == "DeltaTempMean" |
           parameter == "ws" | 
           parameter == "solar" | 
           parameter == "temp" | 
           parameter == "discharge") %>% 
  ggplot(aes(x = as.factor(year), 
             y = value)) + 
  geom_boxplot(aes(color = as.factor(year)))+
  scale_color_manual(values = mb)+
  theme_bw()+
  theme(text = element_text(size=14),
        legend.title=element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 12), 
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        strip.text = element_text(size = 12))+
  facet_wrap(~parameter, scales = "free")
##################################################

#st albans bay 
abAll17_long <- abAll17 %>% 
  mutate(year = year(date), 
         site = rep("StAB")) %>% 
  gather(key = "parameter", value = "value", c(TempMean:discharge))

abAll18_long <- abAll18 %>% 
  mutate(year = year(date), 
         site = rep("StAB")) %>% 
  gather(key = "parameter", value = "value", c(TempMean:discharge))

abAll <- rbind(abAll17_long, abAll18_long)
abAll$doy <- strftime(abAll$date, format = "%j", tz = "Etc/GMT-4")
ab <- c(brewer.pal(3, "Paired")[c(1,2, 3)])


##########--------time series --------##########
abAll %>% 
  filter(parameter == "PCMean1" | 
           parameter == "SpCondMean" | 
           parameter == "pHMean" | 
           parameter == "ODOMean" | 
           parameter == "ChlMean" | 
           parameter == "DeltaTempMean" |
           parameter == "ws" | 
           parameter == "solar" | 
           parameter == "temp" | 
           parameter == "discharge") %>% 
  ggplot(aes(x = doy, 
             y = value, 
             color = as.factor(year),
             group = as.factor(year))) + 
  geom_line()+
  scale_color_manual(values = ab)+
  theme_bw()+
  theme(text = element_text(size=14),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank())+
  facet_wrap(~parameter, scales = "free_y") +
  scale_x_discrete(breaks = seq(120,310,30)) +
  xlab("Day of year")+
  labs(color = "Year")
##################################################


##########--------boxplots--------##########
abAll %>% 
  filter(parameter == "PCMean1" | 
           parameter == "SpCondMean" | 
           parameter == "pHMean" | 
           parameter == "ODOMean" | 
           parameter == "ChlMean" | 
           parameter == "DeltaTempMean" |
           parameter == "ws" | 
           parameter == "solar" | 
           parameter == "temp" | 
           parameter == "discharge") %>% 
  ggplot(aes(x = as.factor(year), 
             y = value)) + 
  geom_boxplot(aes(color = as.factor(year)))+
  scale_color_manual(values = ab)+
  theme_bw()+
  theme(text = element_text(size=14),
        legend.title=element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 12), 
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        strip.text = element_text(size = 12))+
  facet_wrap(~parameter, scales = "free")
##################################################

```