---
title: "Daily Time Lags"
author: "Mahalia Clark"
date: "12/2/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Calculating Daily Averages

First I read in the data: dataframes of hourly buoy data. I then created a function which that takes our hourly data and returns a dataframe of daily averages. It also appends columns with the PC averages 0-6 days into the future (so that all the environmental variables are lagged). Finally, I ran the function on our data so we could use daily averages for subsequent analyses.

```{r daily, warning = FALSE, message = FALSE}
# Load required packages
library(dplyr)
library(ggplot2)
#install.packages("openair")
#install.packages("PerformanceAnalytics")
library(PerformanceAnalytics)
library(openair)

# Load data
MB17 <- read.csv("data/AllDepthData/MB17_allDepths.csv", header = T)
MB18 <- read.csv("data/AllDepthData/MB18_allDepths.csv", header = T)
SA17 <- read.csv("data/AllDepthData/SA17_allDepths.csv", header = T)
SA18 <- read.csv("data/AllDepthData/SA18_allDepths.csv", header = T)

# Function that takes a df of hourly data and returns a df of daily averages,
# including columns with daily PC 0-6 days into the future (foolishly named PC1-7).
dailify <- function(df_in, time_header){
  df_out <- df_in %>%
    select(c(time_header, TEMP_0.5m, SPCOND_0.5m, PH_0.5m, ODO_CONC_0.5m, 
             CHL_RFU_0.5m, TURB_0.5m, deltaDO, deltaTemp, PC_RFU_0.5m)) %>%
    rename(Timestamp = time_header,
           Temp = TEMP_0.5m, 
           SpCond = SPCOND_0.5m, 
           pH = PH_0.5m,
           ODO = ODO_CONC_0.5m,
           Chl = CHL_RFU_0.5m,
           Turb = TURB_0.5m,
           DeltaDO = deltaDO,
           DeltaTemp = deltaTemp,
           PC = PC_RFU_0.5m) %>%
    mutate(timestamp = as.POSIXct(strptime(Timestamp, format = "%m/%d/%Y %H:%M", tz = "Etc/GMT-4")),
           date = as.Date(timestamp, tz = "Etc/GMT-4"), 
           PC = replace(PC, which(PC<0), 0), 
           Chl = replace(Chl, which(Chl<0), 0)) %>% 
    group_by(date) %>%
    summarize(TempMean = mean(Temp, na.rm = F), 
              SpCondMean = mean(SpCond, na.rm = F), 
              pHMean = mean(pH, na.rm = F),
              ODOMean = mean(ODO, na.rm = F),
              ChlMean = mean(Chl, na.rm = F), 
              TurbMean = mean(Turb, na.rm = F),
              DeltaDOMean = mean(DeltaDO, na.rm = F), 
              DeltaTempMean = mean(DeltaTemp, na.rm = F), 
              PCMean1 = mean(PC, na.rm = F))
  df_out$PCMean2 <- lead(df_out$PCMean1)
  df_out$PCMean3 <- lead(df_out$PCMean1, 2)
  df_out$PCMean4 <- lead(df_out$PCMean1, 3)
  df_out$PCMean5 <- lead(df_out$PCMean1, 4)
  df_out$PCMean6 <- lead(df_out$PCMean1, 5)
  df_out$PCMean7 <- lead(df_out$PCMean1, 6)
  return(df_out)
}

# Dailify our data
MB17_daily <- dailify(MB17, "TIMESTAMP")
write.csv(MB17_daily, "MB17_daily.csv")

MB18_daily <- dailify(MB18, "TIMESTAMP")
write.csv(MB18_daily, "MB18_daily.csv")

SA17_daily <- dailify(SA17, "timestamp")
write.csv(SA17_daily, "SA17_daily.csv")

SA18_daily <- dailify(SA18, "timestamp")
write.csv(SA18_daily, "SA18_daily.csv")

```

## Plotting the data

Next, I made a function that creates (messy) timeplots of the data so we can just quickly visually inspect how PC varies with different environmental factors over time.

```{r}
#Function that creates a timeplot of the df (with unlagged PC)
plotify <- function(df){
  timePlot(df, pollutant = c("PCMean1",
                                     "TempMean", 
                                     "SpCondMean",
                                     "ChlMean",
                                     "pHMean", 
                                     "ODOMean",
                                     "TurbMean",
                                     "DeltaDOMean",
                                     "DeltaTempMean"),
           y.relation = "free")
}

# Plot the data
plotify(MB17_daily)
plotify(MB18_daily)
plotify(SA17_daily)
plotify(SA18_daily)
```

## Investigate Correlations

Finally, we were interested in investigating correlations between various environmental factors and different time lags of PC. These tables are a little too crowded, so I will focus on looking at correlations between PC and individual environmental variables that were most important in the Feature Importance results.

```{r}
# Look at all correlations
chart.Correlation(MB17_daily[,2:16], histogram = T)
chart.Correlation(MB18_daily[,2:16], histogram = T)
chart.Correlation(SA17_daily[,2:16], histogram = T)
chart.Correlation(SA18_daily[,2:16], histogram = T)
```

