---
title: "Feature Importance"
author: "WB"
date: "12/3/2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

*going to use Chege's original feature importance code and try and figure out a way to loop the analysis through the 0 - 6 day into the future  data so that we don't have to copy and paste as much 

***what i want to do:** 
 + run a feature importance model for each bay and each year with PC daily mean as the predictor variable 0 - 6 days out 
  + right now Chege has the code to do it one at a time but i want to add a column to each yearly dataset that has DayinFuture so then i can loop through the different predictor varaibles
  + create an empty dataset that has the columns of Site (MB or AB), Year (2017 or 2018), DayinFuture (0-6), Parameter (Temp, ws, wd, ODO...etc.), FeatureImportance (0 - 1.0)
  + right for loop that takes each dataset and runs feature importance for each of the PC daily means and then adds it to the table
  
###Preliminaries 

```{r, warning = F, message = F}
# libraries we're going to use
library(xgboost) # for xgboost
library(tidyverse) # general utility functions
library(data.table)
library(dplyr)
library(Matrix)
library(zoo)
library(Ckmeans.1d.dp)
library(RColorBrewer)

```

### Load in data 

```{r, message = F, warning = F}
###############################################################
# FUNCTION: multMerge
# creates: easy way to upload 4 sep files 
# input: filepath for where data files are located
# output: data frame with all four datasets 
#--------------------------------------------------------------
multmerge <- function(mypath){
  filenames=list.files(path=mypath, full.names=TRUE)
  datalist = lapply(filenames, function(x){read.csv(file=x,header=T)})
  Reduce(function(x,y) {merge(x,y,all = TRUE)}, datalist)
}
###############################################################

data <- multmerge("data/")

# take out variables I don't want to use 
# right now only including the non-correlated variables because we want to learn more about the mechanisms 
dbFull <- data %>% 
  select(-c(X, date, pHMean, TempMean, ODOMean, ChlMean, TurbMean, DeltaDOMean)) %>% 
  mutate(year = as.factor(year)) 

#put DayinFuture into long format so that we can loop through and store the results of the feature importance analysis for easy plotting 
db <- dbFull %>% 
  gather(key = "DayinFuture", value = "PCMean", PCMean1:PCMean7) %>% 
  mutate(DayinFuture = as.numeric(str_sub(DayinFuture, -1))) %>% 
  na.omit()





```


### Create function for feature importance model 

```{r, }

###############################################################
# FUNCTION: featImportance
# creates: runs featImportance code Chege wrote 
# input: data frame with columns SpCondMean, DeltaTempMean, ws, wd, solar, temp, discharge, bay, year, DayinFuture, PCMean
# output: data frame with 5 columns Site, Year, DayinFuture, Parameter, FeatureImportance 
#--------------------------------------------------------------
featImportance <- function(data = db1){
  
  # get training labels i.e our bloom variable
  bloom_variable <- data %>%
    select(PCMean)  # get the column with the bloom measurements
  
  
  #Just have a dataframe with the target variable and lags removed
  without_bloom_indicator <- data %>%
    select(-starts_with("PCMean"))
  
  #Here i remove redundant columns and keep away whatever is not numeric
  without_bloom_numeric <- without_bloom_indicator %>%
     select_if(is.numeric) # select remaining numeric columns

  #convert the dataframe into a matrix
  without_bloom_matrix <- data.matrix(without_bloom_numeric)
  
  # get the numb 70/30 training test split
  numberOfTrainingSamples <- round(nrow(bloom_variable)*.7)
  
  # training data
  train_data <- without_bloom_matrix[1:numberOfTrainingSamples,]
  train_labels <- bloom_variable[1:numberOfTrainingSamples,]
  
 
  # testing data
  test_data <- without_bloom_matrix[-(1:numberOfTrainingSamples),]
  test_labels <- bloom_variable[-(1:numberOfTrainingSamples),]
  
  
  # put our testing & training data into two seperates Dmatrixs objects
  dtrain <- xgb.DMatrix(data = train_data, label= train_labels)
  dtest <- xgb.DMatrix(data = test_data, label= test_labels)
  
  # train a model using our training data
  model <- xgboost(data = dtrain, # the data   
                   nround = 2, # max number of boosting iterations
                   objective = "reg:linear")  
                   # the objective function
  
  # generate predictions for our held-out testing data
  pred <- predict(model, dtest)
  
  # get information on how important each feature is
  importance <- xgb.importance(names(without_bloom_matrix), model = model)
  
  results <- data.frame(list(Parameter = importance$Feature, 
                   FeatureImportance = importance$Gain))
  
  # create output dataframe 
  n <- nrow(results)
  results <- results %>% 
    mutate(bay = rep(data$bay[1], n),
           year = rep(data$year[1], n), 
           DayinFuture = rep(data$DayinFuture[1]-1, n))
  return(results)
  
}

###############################################################

#useful code to see the unused parameters that got dropped 
#unused_features <- setdiff(colnames(dtrain), importance$Feature)

```


### 1. Separate by year and bay 

```{r, include = F}
# splitting them up by bay because i want to run feature importance for each bay and year 
db1 <- db %>%
  filter(bay == "MB" & year == "2017")

db2 <- db %>%
  filter(bay == "MB" & year == "2018") 

db3 <- db %>%
  filter(bay == "AB" & year == "2017")

db4 <- db %>%
  filter(bay == "AB" & year == "2018")

#necessary to run these first so i acn get oproper utputs for loops i will run 
myList=NULL
df=NULL
datalist = list()

######################MB 2017##############################
for (i in 1:7) {
  temp <- db1[(db1$DayinFuture == i),]
  df <- featImportance(temp)
  datalist[[i]] <- df
  
}

mb17 <- do.call(rbind, datalist)
###############################################################



######################MB 2018##############################
for (i in 1:7) {
  temp <- db2[(db2$DayinFuture == i),]
  df <- featImportance(temp)
  datalist[[i]] <- df
  
}

mb18 <- do.call(rbind, datalist)
###############################################################


######################AB 2017##############################
for (i in 1:7) {
  temp <- db3[(db3$DayinFuture == i),]
  df <- featImportance(temp)
  datalist[[i]] <- df
  
}

ab17 <- do.call(rbind, datalist)
###############################################################


######################AB 2018##############################
for (i in 1:7) {
  temp <- db4[(db4$DayinFuture == i),]
  df <- featImportance(temp)
  datalist[[i]] <- df
  
}

ab18 <- do.call(rbind, datalist)
###############################################################



```

#### 1.1 Plot: stacked bar chart - separated by bay and year 

```{r, }

#combine outputs into one dataset to make plot 
results_1 <- rbind(ab17, ab18, mb17, mb18)

#write.csv(results_1, "Sep_bay_and_year.csv")

results_1 %>% 
  ggplot() +
  geom_bar(aes(x = as.factor(DayinFuture), y = FeatureImportance, fill = Parameter),
           stat = 'identity')+
  scale_y_continuous(limits = c(0, 1),
                     name = "Feature importance") + 
  scale_fill_brewer(palette = "Paired", labels = c(Delta~"Water temperature", "Discharge", "Solar radiation", "Specific conductivity", "Air temperature", "Wind direction", "Wind speed"))+
  xlab("Days in the future")+
  #expand = c(0,0))+
  theme_bw()+
  theme(
        axis.title = element_text(size = 12),
        axis.text = element_text(size =10, colour = "black"),
        legend.text = element_text(size = 10),
        legend.text.align = 0,
        axis.text.x = element_text(angle = 45, hjust = 1))+
  facet_wrap(~bay+year)

```


### 2. Keep all data together - combined bay and year  

- to-do: figure how to run chege's code with categorical variables, check the old, original code chege uploaded ! 
- i figured out the function i wrote wasn't working with the bays and years combined because i had changed them to categorical varaibles for the above analysis to separate them and make 4 plots
- now i need to go back and change "bay" and "year" to 0's and 1's: 2017 = 0, 2018 = 1, MB = 0, "AB" = 1. 

```{r, }
#first need to convert the categorical variables back into numeric
#want to keep the original column to make sure it's correct
db$bayNUM <- db$bay
levels(db$bayNUM) <- c(0,1) #0 is MB, 1 is AB
#now for year
db$yearNUM <- db$year
levels(db$yearNUM) <- c(0, 1) #0 is 2017, 1 is 2018

#looks good, now i need to make them numbers and take out the categorical variables
dbALL <- db %>%
  mutate(bayNUM = as.numeric(bayNUM),
         yearNUM = as.numeric(yearNUM))  %>%
  select(-c(bay, year))
#weird i automatically changed 0 and 1 to 1 and 2, not sure if that matters

#i thought i needed to create a sparse matrix for the categorical variables to work but i'm 99% sure the results are exactly the same...here's the code but i didn't work it into the function because the test run i did with just one day of data was the same 

# test <- db %>% 
#   filter(DayinFuture == 1) %>% 
#   select(-DayinFuture)
# 
# 
# sparse_matrix <- sparse.model.matrix(PCMean ~ ., 
#                                      data = test,
#                                      contrasts.arg=list(bay=diag(nlevels(db$bay)), 
#             year=diag(nlevels(db$year))))[,-1]
# head(sparse_matrix)


myList=NULL
df=NULL
datalist = list()
######################all data ##############################
for (i in 1:7) {
  temp <- dbALL[(dbALL$DayinFuture == i),]
  df <- featImportance(temp)
  datalist[[i]] <- df

}


###############################################################



```


#### 2.1 Plot: stacked bar chart - all data combined  

```{r, }

#combine outputs into one dataset to make plot 
results_2 <- do.call(rbind, datalist)
#write.csv(results_2, "allCombined.csv")

results_2 %>% 
  ggplot() +
  geom_bar(aes(x = as.factor(DayinFuture), y = FeatureImportance, fill = Parameter),
           stat = 'identity') +
  scale_y_continuous(limits = c(0, 1),
                     name = "Feature importance") + 
  scale_fill_brewer(palette = "Paired", labels = c(Delta~"Water temperature", "Discharge", "Solar radiation", "Specific conductivity", "Air temperature", "Wind direction", "Wind speed", "Year"))+
  xlab("Days in the future")+
  #expand = c(0,0))+
  theme_bw()+
  theme(
        axis.title = element_text(size = 12),
        axis.text = element_text(size =10, colour = "black"),
        legend.text = element_text(size = 10),
        legend.text.align = 0,
        axis.text.x = element_text(angle = 45, hjust = 1))

```


### 3. Separate by bay

```{r, include = F}
#need to add back bay into categorical variable 
#first need to convert the categorical variables back into numeric
#want to keep the original column to make sure it's correct
#now for year
db$yearNUM <- db$year
levels(db$yearNUM) <- c(0, 1) #0 is 2017, 1 is 2018

#looks good, now i need to make them numbers and take out the categorical variables
dbBAY<- db %>%
  mutate(
         yearNUM = as.numeric(yearNUM))  %>%
  select(-c(year, bayNUM))

# splitting them up by bay because i want to run feature importance for each bay and year 
db5 <- dbBAY %>%
  filter(bay == "MB")

db6 <- dbBAY %>%
  filter(bay == "AB")

#necessary to run these first so i acn get oproper utputs for loops i will run 
myList=NULL
df=NULL
datalist = list()

###################### MB ##############################
for (i in 1:7) {
  temp <- db5[(db5$DayinFuture == i),]
  df <- featImportance(temp)
  datalist[[i]] <- df
  
}

mb <- do.call(rbind, datalist)
###############################################################



###################### AB ##############################
for (i in 1:7) {
  temp <- db6[(db6$DayinFuture == i),]
  df <- featImportance(temp)
  datalist[[i]] <- df
  
}

ab <- do.call(rbind, datalist)
###############################################################



```


#### 3.1 Plot: stacked bar chart - separated by bay 

```{r, }

#combine outputs into one dataset to make plot 
results_3 <- rbind(ab, mb)

write.csv(results_3, "Sep_bay.csv")

results_3 %>% 
  ggplot() +
  geom_bar(aes(x = as.factor(DayinFuture), y = FeatureImportance, fill = Parameter),
           stat = 'identity') +
  scale_y_continuous(limits = c(0, 1),
                     name = "Feature importance") + 
  scale_fill_brewer(palette = "Paired", labels = c(Delta~"Water temperature", "Discharge", "Solar radiation", "Specific conductivity", "Air temperature", "Wind direction", "Wind speed", "Year"))+
  xlab("Days in the future")+
  #expand = c(0,0))+
  theme_bw()+
  theme(
        axis.title = element_text(size = 12),
        axis.text = element_text(size =10, colour = "black"),
        legend.text = element_text(size = 10),
        legend.text.align = 0,
        axis.text.x = element_text(angle = 45, hjust = 1))+
  facet_wrap(~bay, nrow = 2)

```

### 4. Separate by year

```{r, include = F}
#need to add back bay into categorical variable 
#first need to convert the categorical variables back into numeric
#want to keep the original column to make sure it's correct
#now for bay
db$bayNUM <- db$bay
levels(db$bayNUM) <- c(0, 1) #0 is 2017, 1 is 2018

#looks good, now i need to make them numbers and take out the categorical variables
dbYR <- db %>%
  mutate(
         bayNUM = as.numeric(bayNUM))  %>%
  select(-c(bay, yearNUM))

# splitting them up by bay because i want to run feature importance for each bay and year 
db7 <- dbYR %>%
  filter(year == "2017")

db8 <- dbYR %>%
  filter(year == "2018") 

#necessary to run these first so i acn get oproper utputs for loops i will run 
myList=NULL
df=NULL
datalist = list()

###################### MB ##############################
for (i in 1:7) {
  temp <- db7[(db7$DayinFuture == i),]
  df <- featImportance(temp)
  datalist[[i]] <- df
  
}

db17 <- do.call(rbind, datalist)
###############################################################



###################### AB ##############################
for (i in 1:7) {
  temp <- db8[(db8$DayinFuture == i),]
  df <- featImportance(temp)
  datalist[[i]] <- df
  
}

db18 <- do.call(rbind, datalist)
###############################################################



```


#### 4.1 Plot: stacked bar chart - separated by year 

```{r, }

#combine outputs into one dataset to make plot 
results_4 <- rbind(db17, db18)

write.csv(results_4, "Sep_year.csv")

results_4 %>% 
  ggplot() +
  geom_bar(aes(x = as.factor(DayinFuture), y = FeatureImportance, fill = Parameter),
           stat = 'identity') +
  scale_y_continuous(limits = c(0, 1),
                     name = "Feature importance") + 
  scale_fill_brewer(palette = "Paired", labels = c(Delta~"Water temperature", "Discharge", "Solar radiation", "Specific conductivity", "Air temperature", "Wind direction", "Wind speed"))+
  xlab("Days in the future")+
  #expand = c(0,0))+
  theme_bw()+
  theme(
        axis.title = element_text(size = 12),
        axis.text = element_text(size =10, colour = "black"),
        legend.text = element_text(size = 10),
        legend.text.align = 0,
        axis.text.x = element_text(angle = 45, hjust = 1))+
  facet_wrap(~year, nrow = 2)

```
