---
title: "ListofParams"
author: "WB"
date: "11/21/2019"
output:
  pdf_document: default
  html_document: default
---

# Figuring out which parameters we want to include for the feature importance tests

### Preliminaries 

```{r, message=FALSE, warning = FALSE}
#load the column headers
params <-  c("TempMean", 
             "SpCondMean", 
             "pHMean", 
             "ODOMean", 
             "ChlMean", 
             "TurbMean", 
             "DeltaDOMean", 
             "DeltaTempMean",
             "PCMean1", 
             "PCMean2",
             "PCMean3",
             "PCMean4", 
             "PCMean5", 
             "PCMean6",
             "PCMean7",
             "ws",
             "wd",
             "solar",
             "temp",
             "discharge")
```

### Set 1  

```{r, message=FALSE, warning = FALSE}

set1_0 <- c(params[2:5], params[8:9], params[16:20])
set1_1 <- c(params[2:5], params[8], params[10], params[16:20])
set1_2 <- c(params[2:5], params[8], params[11], params[16:20])
set1_3 <- c(params[2:5], params[8], params[12], params[16:20])
set1_4 <- c(params[2:5], params[8], params[13], params[16:20])
set1_5 <- c(params[2:5], params[8], params[14], params[16:20])
set1_6 <- c(params[2:5], params[8], params[15], params[16:20])

```


**Set 1 includes:** *`r params[2:5]`*, *`r params[8]`*, *`r params[16:20]`*

**Set 1 does not include:**

- *`r params[1]`* (water temp) --> going to use air temp instead
- *`r params[6]`* (turbidity) --> too closely correlated with PCMeans
- *`r params[7]`* (dissolved oxygen stratification) --> going to use deltaTempMean instead which is a better proxy for stratification (DeltaDO is affected by biological activity too much)

The subsequent runs in **Set 1** have the lag times incorporated for PC: 

- Predicting PC for that day (Set1_0): *`r set1_0`*
- Predicting PC 1 day out (Set1_1): *`r set1_1`*
- Predicting PC 2 days out (Set1_2): *`r set1_2`*
- Predicting PC 3 days out (Set1_3): *`r set1_3`*
- Predicting PC 4 days out (Set1_4): *`r set1_4`*
- Predicting PC 5 days out (Set1_5): *`r set1_5`*
- Predicting PC 6 days out (Set1_6): *`r set1_6`*

### Set 2

```{r, message=FALSE, warning = FALSE}

set2_0 <- c(params[3:5], params[8:9], params[16], params[18:20])
set2_1 <- c(params[3:5], params[8], params[10], params[16], params[18:20])
set2_2 <- c(params[3:5], params[8], params[11], params[16], params[18:20])
set2_3 <- c(params[3:5], params[8], params[12], params[16], params[18:20])
set2_4 <- c(params[3:5], params[8], params[13], params[16], params[18:20])
set2_5 <- c(params[3:5], params[8], params[14], params[16], params[18:20])
set2_6 <- c(params[3:5], params[8], params[15], params[16], params[18:20])

```


**Set 2 includes:** *`r params[3:5]`*, *`r params[8]`*, *`r params[16]`*, *`r params[18:20]`*

**Set 2 does not include:**

- *`r params[2]`* (specific conductivity) --> i don't think this actually affects PC and it might be giving a false importance signal
- *`r params[17]`* (wind direction) --> i'm not sure it's useful to have wind direction in as a continuous variable so it might cloud other relationships...
- *`r params[1]`*(water temp) --> going to use air temp instead
- *`r params[6]`* (turbidity) --> too closely correlated with PCMeans
- *`r params[7]`* (dissolved oxygen stratification) --> going to use deltaTempMean instead which is a better proxy for stratification (DeltaDO is affected by biological activity too much)


Runs in **Set 2** have the lag times incorporated for PC: 

- Predicting PC for that day (Set2_0): *`r set2_0`*
- Predicting PC 1 day out (Set1_1): *`r set2_1`*
- Predicting PC 2 days out (Set1_2): *`r set2_2`*
- Predicting PC 3 days out (Set1_3): *`r set2_3`*
- Predicting PC 4 days out (Set1_4): *`r set2_4`*
- Predicting PC 5 days out (Set1_5): *`r set2_5`*
- Predicting PC 6 days out (Set1_6): *`r set2_6`*