---
title: "Which Method makes the Prettiest Time Plots"
author: "Mahalia Clark"
date: "12/4/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Time Plots

We wanted to make some prettier time plots to better visualize the data. I tried three methods: basic built in plot function, timePlot function from openair package, and ggplot.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#Load relevant packages
library(dplyr)
library(tidyr)
library(openair)
library(ggplot2)
library(viridis)

#install.packages("padr")
  library(padr)
  library(zoo)

#Load Data
MB17 <- read.csv("data/mbAll17.csv", header = T) 
MB18 <- read.csv("data/mbAll18.csv", header = T)
SA17 <- read.csv("data/abAll17.csv", header = T)
SA18 <- read.csv("data/abAll18.csv", header = T)

#Fix Date Format
MB17 <- MB17 %>% 
  mutate(date = as.POSIXct(strptime(date, format = "%m/%d/%Y", tz = "Etc/GMT-4")))
MB17 <- pad(MB17, interval = "day")

MB18 <- MB18 %>% 
  mutate(date = as.POSIXct(strptime(date, format = "%m/%d/%Y", tz = "Etc/GMT-4")))
MB18 <- pad(MB18, interval = "day")

SA17 <- SA17 %>% 
  mutate(date = as.POSIXct(strptime(date, format = "%m/%d/%Y", tz = "Etc/GMT-4")))
SA17 <- pad(SA17, interval = "day")

SA18 <- SA18 %>% 
  mutate(date = as.POSIXct(strptime(date, format = "%m/%d/%Y", tz = "Etc/GMT-4")))
SA18 <- pad(SA18, interval = "day")

```

## First we can try with the basic plot function

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 6, fig.width = 7}
plotf <- function(df, title){
  par(mfrow = c(8, 1), oma = c(5,0,0,0) + 0.1, mar = c(0,5,1,0) + 0.1)
  plot(MB17$date, MB17$PCMean1, type = "l", main = title, ylab = "PC", xaxt = "n")
  plot(MB17$date, MB17$DeltaTempMean, type = "l", ylab = "Strat", xaxt = "n")
  plot(MB17$date, MB17$discharge, type = "l", ylab = "Q", xaxt = "n")
  plot(MB17$date, MB17$solar, type = "l", ylab = "Sun", xaxt = "n")
  plot(MB17$date, MB17$SpCondMean, type = "l", ylab = "Sal", xaxt = "n")
  plot(MB17$date, MB17$temp, type = "l", ylab = "Temp", xaxt = "n")
  plot(MB17$date, MB17$wd, type = "l", ylab = "WD", xaxt = "n")
  plot(MB17$date, MB17$ws, type = "l", ylab = "WS")
}
plotf(MB17, "MB17")
```

The graphs are a little short vertically. How can I get rid of all the space in between without losing my bottom x axis?

## Next, we can try using the package Wilton suggested: openair.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 6, fig.width = 7}
#Try with timePlot
plotify <- function(df, title){
    timePlot(df, pollutant = c("PCMean1",
                                     "DeltaTempMean",
                                     "discharge",
                                     "SpCondMean",
                                     "temp",
                                     "solar",
                                     "ws",
                                     "wd"),
           date.pad = TRUE,
           key = FALSE,
           name.pol = c("PC", "Strat", "Q", "Sal", "Temp", "Sun", "WS", "WD"),
           ylab = " ",
           main = title,
           y.relation = "free") 
           
}
plotify(MB17, "MB17")
#plotify(MB18, "MB18")
#plotify(SA17, "SA17")
#plotify(SA18, "SA18")

```
This is pretty good. I just feel like the y axes look messy.

## Finally, we can try using ggplot.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 6, fig.width = 7}
#Select&Rename
shorten <- function(df){
  newdf <- df %>% 
    select(c(date, "PCMean1",
                                     "DeltaTempMean",
                                     "discharge",
                                     "SpCondMean",
                                     "temp",
                                     "solar",
                                     "ws",
                                     "wd")) %>% 
    rename(PC = PCMean1,
         Strat = DeltaTempMean,
         Q = discharge,
         Sal = SpCondMean,
         Temp = temp,
         Sun = solar,
         WS = ws,
         WD = wd)
    return(newdf)
}

MB17_short <- shorten(MB17)
MB17_short$day <- strftime(MB17_short$date, format = "%j", tz = "Etc/GMT-4")

MB18_short <- shorten(MB18)
MB18_short$day <- strftime(MB18_short$date, format = "%j", tz = "Etc/GMT-4")


#Tidy
MB17_tidy <- gather(MB17_short, "PC":"WD", key = "Variable", value = "Value") 
MB18_tidy <- gather(MB18_short, "PC":"WD", key = "Variable", value = "Value") 

#Plot
#ggplot(MB17_tidy, aes(x = date, y = Value, group = Variable, color = Variable)) + geom_line(show.legend = FALSE) + labs(title = "MB17") + xlab("Time") + theme(axis.title.y = element_blank())  + facet_grid(rows = "Variable", scales = "free_y", switch = "both") 

#Trying to combine years onto one plot...
ggplot(NULL, aes(x = day, y = Value, group = Variable, color = Variable)) +
  geom_line(data = MB17_tidy, show.legend = FALSE) + 
  geom_line(data = MB18_tidy, show.legend = FALSE, linetype = "dashed") + 
  labs(title = "MB") + xlab("Day of year") + 
  facet_grid(rows = "Variable", scales = "free_y", switch = "both") +
  scale_x_discrete(breaks = seq(120,310,30))+
  theme_light() 

#Woah, it worked!! But can I move the labels to the right? Yes! (switch)
```
It's beautiful!
