---
title: "Stacked Bar Graph"
author: "WB"
date: "12/2/2019"
output: html_document
---

#### Code from previous project where i made a stacked bar graph: 

```{r, } 

# how i organized the data frame,  not sure if helpful 
db2 <- db %>% 
  group_by(Date, Site, Genus, Vo, Vc, nFOV, FOVmL, Year) %>% 
  summarize(nCells = sum(nCells)) %>% 
  mutate(cellsFOV = nCells/nFOV,
         cellsmLConc = cellsFOV/FOVmL, 
         cellsmL = ((cellsmLConc*Vc)/Vo))
#making the plot 
db2 %>% 
  ggplot() +
  geom_bar(aes(x = as.factor(Date), y = cellsmL, fill = Names),
           stat = 'identity') +
  scale_y_continuous(limits = c(0, 200000),name = expression("Phytoplankton"~(cells~mL^-1)))+ #expand = c(0,0))+
  theme_bw()+
  theme(axis.title.x = element_blank(),
        axis.title = element_text(size = 12),
        axis.text = element_text(size =10, colour = "black"),
        legend.text = element_text(size = 10),
        legend.text.align = 0,
        axis.text.x = element_text(angle = 45, hjust = 1))+
  facet_wrap(~Site+Year)

```


### Now ammend the code to match my data 

- I want the x-axis to be "Days in the future" and the y-axis to be "Feature importance" and there will be 7 stacked bar plots that add up to 1 for days 0 - 6 in the future with all the relative importance of the parameters 
- want 4 plots (one for each year and bay)
- so input data frame should be one data frame with the colomns of Site, Year, DayInFuture, Parameter, FeatureImportance

```{r, }

all %>% 
  ggplot() +
  geom_bar(aes(x = as.factor(DayinFuture), y = FeatureImportance, fill = Parameter),
           stat = 'identity') +
  scale_y_continuous(limits = c(0, 1),
                     name = "Feature importance") + 
  xlab("Days in the future")+
  #expand = c(0,0))+
  theme_bw()+
  theme(
        axis.title = element_text(size = 12),
        axis.text = element_text(size =10, colour = "black"),
        legend.text = element_text(size = 10),
        legend.text.align = 0,
        axis.text.x = element_text(angle = 45, hjust = 1))+
  facet_wrap(~bay+year)

```
